---
title: "compphysiol: Computational Physiology toolbox"
format: gfm
editor: visual
---

<!-- badges: start -->

[![R-CMD-check](https://github.com/niklhart/compphysiol-R/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/niklhart/compphysiol-R/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/niklhart/compphysiol-R/graph/badge.svg)](https://app.codecov.io/gh/niklhart/compphysiol-R)

<!-- badges: end -->

The Computational Physiology Toolbox is an R package for computations with physiology-based models, in particular physiologically-based pharmacokinetic (PBPK) models. It features a modular object-oriented design and an intuitive scripting language.

## Installation

You can download the development version of the package on GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("niklhart/compphysiol-R")
```

## Usage - PBPK modeling workflow

```{r}
library(compphysiol)
```

For demonstrating a PBPK workflow, we will use the `sMD_PBPK_12CMT_wellstirred()` model, which is a generic 12-compartment whole-body PBPK model with well-stirred tissue compartments. We will load physiological parameters for a 35-year-old human and drug parameters for Warfarin from the built-in databases. We will then calculate tissue partition coefficients using the Rodgers and Rowland method.

```{r}
M <- sMD_PBPK_12CMT_wellstirred()

phys <- loadphysiology("human35m")
drug <- loaddrugdata("Warfarin", species = "human")

compnames <- names(M$compartments)
organs <- setdiff(compnames, c("ven","art"))
Ktis <- rodgersrowland(phys, drug, organs = organs)
```

Physiological parameters (e.g., tissue volumes and blood flows), drug parameters (e.g., blood-to-plasma ratio and clearance) and tissue partition coefficients can be added to the model as follows:

```{r}
param <- c(
    param(phys, name = c("Vtis","Qblo"), organs = organs),
    param(drug, name = c("BP","CL")),
    Ktis
)
M <- add_parameter(M, param = param)
```

Add a dosing regimen:

```{r}
M <- add_dosing(M, target = "ven", time = 0 [h], amount = 100 [mg])
```

Convert to ODEs and simulate:

```{r}
odeinfo <- to_ode(M, dimensions = list(time = "h", mass = "mg", length = "cm"))
times <- seq(0, 24, by = 0.1)
out <- deSolve::ode(
    y = odeinfo$y0,
    times = times,
    func = odeinfo$odefun,
    parms = paramValues,
    events = odeinfo$events
)
out_df <- as.data.frame(out)
```

Plot compartment amounts
```{r}
matplot(
    out_df$time,
    out_df[, -1],
    type = "l",
    lty = 1,
    col = 1:12,
    xlab = "Time (h)",
    ylab = "Amount",
    main = "PBPK Model Compartments"
)
legend(
    "topright",
    legend = colnames(out_df)[-1],
    col = 1:12,
    lty = 1
)
```

Evaluate observables at each time point

```{r}
obs_df <- as.data.frame(sapply(
    odeinfo$obsFuncs,
    function(f) f(out_df$time, as.matrix(out_df[, -1]), paramValues)
))
names(obs_df) <- names(odeinfo$obsFuncs)
obs_df$time <- out_df$time
```

Plot plasma concentration

```{r}
plot(
    obs_df$time,
    obs_df$Cpla,
    type = "l",
    col = "black",
    lwd = 2,
    xlab = "Time (h)",
    ylab = "Plasma concentration"
)
```