---
title: "compphysiol: Computational physiology toolbox"
format: gfm
editor: visual
---

<!-- badges: start -->

[![R-CMD-check](https://github.com/niklhart/compphysiol-R/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/niklhart/compphysiol-R/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/niklhart/compphysiol-R/graph/badge.svg)](https://app.codecov.io/gh/niklhart/compphysiol-R)

<!-- badges: end -->

`compphysiol` is an R package for computations with physiology-based models, in particular physiologically-based pharmacokinetic (PBPK) models. It features a modular object-oriented design and an intuitive scripting language.

## Installation

You can download the development version of the package on GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("niklhart/compphysiol-R")
```

## Citation

See `citation("compphysiol")` for how to cite this package in scientific work.


## Usage

```{r}
library(compphysiol)
```

This example defines and simulates a two-compartment empirical PK model with
unit-aware amounts, volumes, rate constants, and simulation times.

```{r}
model <- multiCompModel(ncomp = 2, type = "micro", unit = "mg") |>
    add_dosing(time = 0 [h], amount = 100 [mg], cmt = "cen") |>
    wire(what = "molec")

pars <- parameters(
    kc0 = 0.15 [1/h],
    kcp = 0.08 [1/h],
    kpc = 0.05 [1/h],
    Vcen = 8 [L],
    Vper = 20 [L]
)
```

> **Note**
> RStudio may currently flag expressions such as `100 [mg]` as syntax errors
> even though they are valid R code and run correctly. This also affects
> ordinary bracket calls outside `compphysiol`, such as `1 [TRUE]`, so it is an
> RStudio diagnostics issue rather than a package error. Current workarounds
> are switching diagnostics off for affected files with `# !diagnostics off`,
> switching diagnostics off globally in RStudio, or using another IDE.

Simulate the model over 24 hours:

```{r}
sim <- simulate(
    model,
    time = seq(0, 24, by = 0.5) [h],
    parameters = pars
)

head(sim$states)
head(sim$observables)
```

The returned state and observable tables keep their units. If `ggplot2`,
`reshape2`, and `units` are installed, they can be plotted directly:

```{r, message = FALSE, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("reshape2", quietly = TRUE) && requireNamespace("units", quietly = TRUE)}
library(ggplot2)
library(units)

states_long <- reshape2::melt(
    sim$states,
    id.vars = "time",
    variable.name = "state",
    value.name = "amount"
)

ggplot(states_long, aes(x = time, y = amount)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(vars(state), scales = "free_y") +
    labs(x = "Time", y = "Amount")
```

## AI Agent Guidance

This repository intentionally includes an AI-agent guidance file `AGENTS.md`
and project-specific Codex skills. These files document repository
conventions, modeling vocabulary, testing expectations, and review workflows
for AI-assisted development.
