---
title: "Estimating a unit-aware model"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Estimating a unit-aware model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates maximum-likelihood estimation with `stats::optim()`.
The workflow keeps four concerns separate:

- the compartment model describes the dynamics;
- experiments hold known parameters, dosing, and observation data;
- the statistical model describes observation variability;
- the estimation engine controls the numerical optimizer.

All physical quantities retain their units at the model and observation
boundaries. Optimization itself uses transformed numeric coordinates.

```{r}
#| label: setup
library(compphysiol)
library(units)
library(ggplot2)
```

## Dynamic model

We use a one-compartment model with an initial amount of 100 mg in a 1 L
central compartment. The first-order elimination rate `k` will be estimated.
The model parameter value is a simulation default; the estimation problem
provides the optimizer's initial estimate separately.

```{r}
#| label: model
model <- compartment_model() |>
    add_compartment("Central", volume = 1 [L]) |>
    add_molecule(
        "drug",
        cmt = "Central",
        initial = 100 [mg],
        type = "amount"
    ) |>
    add_transport(
        from = "Central",
        to = NULL,
        molec = "drug",
        const = "k"
    ) |>
    add_observable(C = c[drug, Central]) |>
    add_parameter(k = 0.2 [1/h])

model
```

## Observation data and experiment

`ObservationData` use the canonical long-format columns `time`, `observable`,
and `value`. Unlike an `ObservationSchedule`, observation data contain values
and can therefore contribute to an estimation likelihood.

```{r}
#| label: observations
observed <- observation_data(
    time = c(1, 2, 4) [h],
    observable = "C",
    value = c(82, 68, 45) [mg/L]
)

study <- experiments(
    individual_1 = experiment(
        observations = observed,
        start = 0 [h]
    )
)

study
```

Each experiment is one independent individual estimation unit. Its collection
name is the individual identifier. An experiment may also contain known
parameters or a dosing schedule. Individual covariates belong in these known
parameters and cannot simultaneously be declared as shared estimated
parameters. Relationships derived from covariates belong in model equations.

## Statistical model

The statistical model is independent of the dynamic model and stores unresolved
distribution specifications. Here, concentration measurements have independent
normal error around the model prediction with standard deviation `sigma`.
Because the observable is a concentration, `sigma` is estimated in concentration
units.

```{r}
#| label: statistical-model
statistics <- statistical_model(
    C = normal(sd = "sigma")
)
```

`normal(sd = proportional("sigma"))` describes proportional error,
`normal(sd = combined(...))` describes combined error, and `lognormal()`
describes log-normal variability. Proportional and log-normal scale parameters
are dimensionless; the constant part of a combined specification has the
observable's units.

## Estimated parameters

Every estimated parameter has an explicit initial estimate. Bounds and an
internal transformation are optional. Log transformations are convenient for
strictly positive rate and scale parameters. Omitted bounds adapt to the
transformation and inherit the initial value's units: identity uses an unbounded
interval, log uses a zero lower bound, and logit uses the interval from zero to one.

```{r}
#| label: estimated-parameters
specs <- c(
    k = parameter_spec(
        initial = 0.1 [1/h],
        transform = "log"
    ),
    sigma = parameter_spec(
        initial = 1 [mg/L],
        transform = "log"
    )
)

specs
```

The transformation affects only the optimizer coordinates. Returned estimates
use the units in which the parameters were declared.

## Estimation problem and engine

An `EstimationProblem` is the engine-neutral scientific specification. Engine
settings are supplied separately, so the same problem can later be evaluated
with another estimation engine.

```{r}
#| label: problem
problem <- estimation_problem(
    model = model,
    experiments = study,
    parameters = specs,
    statistics = statistics
)

engine <- optim_engine(
    method = "L-BFGS-B",
    control = list(maxit = 200)
)
```

## Fit and inspect the result

```{r}
#| label: fit
fit <- estimate(problem, engine = engine)

coef(fit)
fit$neg_log_lik
fit$convergence
```

`coef()` returns a `Parameters` object and therefore retains parameter units.
`neg_log_lik` is the joint negative log-likelihood. Engine-independent
convergence information is available in `fit$convergence`, while the complete
`stats::optim()` return value remains available as `fit$engine_result`.

Fitted values and observed-minus-fitted residuals are `ObservationData` objects.
They retain observation-row order and an experiment identifier as well as
unit-bearing `time` and `value` columns. Rows with missing observations are
retained: their fitted value is available and their residual is missing.

```{r}
#| label: diagnostics
fitted(fit)
residuals(fit)
```

## Goodness-of-fit plot

`fitted(fit)` contains predictions at the observation times. For a smooth
time-course line, simulate the fitted dynamic parameters on a finer observation
schedule. Observation-error parameters such as `sigma` are not needed for this
deterministic prediction.

```{r}
#| label: goodness-of-fit
prediction_grid <- observation_schedule(
    time = seq(0, 4, length.out = 101) [h],
    observable = "C"
)

prediction_experiment <- experiment(
    parameters = coef(fit)["k"],
    observations = prediction_grid,
    start = 0 [h]
)

prediction_curve <- simulate(
    model,
    experiment = prediction_experiment
)$observables

ggplot() +
    geom_line(
        data = prediction_curve,
        aes(x = time, y = value),
        linewidth = 0.8,
        colour = "#3366AA"
    ) +
    geom_point(
        data = observed,
        aes(x = time, y = value),
        size = 2.5,
        colour = "#AA3333"
    ) +
    labs(
        x = "Time",
        y = "Central concentration",
        title = "Observed and fitted concentration-time profile"
    ) +
    theme_minimal()
```

## Reusing a lowered model

Estimation also accepts `ProcessModel`, `OdeModel`, and `CompiledOdeModel`
representations. A compiled model avoids rebuilding its solver-facing structure
during repeated objective evaluations. The estimation specification otherwise
remains unchanged.

```{r}
#| label: compiled-problem
compiled_problem <- estimation_problem(
    model = to_compiled_ode_model(model),
    experiments = study,
    parameters = specs,
    statistics = statistics
)

compiled_fit <- estimate(compiled_problem, engine = engine)
coef(compiled_fit)
```

Analytical and stochastic model representations are intentionally outside the
current estimation contract.
