---
title: "Simulating and plotting unit-aware models"
format:
  html:
    toc: true
    number-sections: false
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Simulating and plotting unit-aware models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette shows a compact workflow from model definition to simulation and
plotting. The model is specified with units, `simulate()` returns unit-aware
state and observable trajectories, and `ggplot2` can plot these columns
directly.

```{r}
#| label: setup
library(compphysiol)
library(units)
library(ggplot2)
library(reshape2)
```

## Model

We use a two-compartment empirical PK model in micro-parameterization. The
integrated states are amounts in the central and peripheral compartments. 
Model parameters are defined with units using `add_parameter()`.
The model also defines the central concentration observable `Ccen`.

```{r}
#| label: model
model <- multiCompModel(ncomp = 2, type = "micro", unit = "mg") |>
    add_dosing(time = 0 [h], amount = 100 [mg], cmt = "cen") |>
    add_parameter(
        kc0 = 0.15 [1/h],
        kcp = 0.08 [1/h],
        kpc = 0.05 [1/h],
        Vcen = 8 [L],
        Vper = 20 [L]
    )
model
```

## Simulation

`simulate()` allows to simulate a compartment model (by default, using ODEs). 
It handles units properly, ensuring that the simulation results are unit-aware.

```{r}
#| label: simulate
sim <- simulate(
    model,
    time = seq(0, 24, by = 0.5) [h]
)

sim
head(sim$states)
head(sim$observables)
```

## Long Format

The state table returned by `simulate()` is wide: one column for time and one
column for each ODE state. For facetted plotting, it is often convenient to
reshape this to long format.

```{r}
#| label: reshape
states_long <- melt(
    sim$states,
    id.vars = "time",
    variable.name = "state",
    value.name = "amount"
)

head(states_long)
```

## Plotting

`ggplot2` handles `units` columns directly. Unit labels are moved to axis
captions automatically, and facets work as usual.

```{r}
#| label: plot-states
ggplot(states_long, aes(x = time, y = amount)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(vars(state), scales = "free_y") +
    labs(x = "Time", y = "Amount")
```
